---
title: "Homework 1 Adv. Data Programming in R"
author: "Venkata Vivek Mudunuru 23200316"
format: html
editor: visual
---

# Dataset - AirPassengers

```{r}
#Working on the dataset of the passengers for flight from the year 1949 to 1960 
dts = (AirPassengers)
```

# Maximum Number of Flights and Plot

```{r}
# Plotting the histogram to understand the trend in which month 
# and which year were the highest number of passengers was observed
# Load necessary packages
library(dplyr)
library(ggplot2)

#Since it is a time series data we need to convert into a dataframe

df <- data.frame(
  year = rep(1949:1960, each = 12), #for all the years mentioned
  month = rep(1:12, times = 12), #and the months for that year
  passengers = as.numeric(dts) 
)

annual_passengers <- df %>% #Now grouping them in annually based on year
  group_by(year) %>%
  summarise(total_passengers = sum(passengers)) #and finding the sum of this accordingly

# Printing the annual passengers data frame
print(annual_passengers)
```

```{r}
# As observed, now filtering the data for the year with the highest total passengers
max_year <- 1960
max_year_data <- df %>% filter(year == max_year)
# Converting the month numbers to month names in the max_year_data data frame
max_year_data$month <- factor(max_year_data$month, levels = 1:12, labels = month.name)

# Accordingly creating a histogram of the monthly passenger counts for the year with the highest total
ggplot(max_year_data, aes(x = month, y = passengers)) +
  geom_col(fill = "lightblue", color = "black") +
  labs(title = paste("Monthly Passengers in the Year with the Highest Total (", max_year, ")", sep = ""),
       x = "Month",
       y = "Number of Passengers") +
  theme_minimal()
```

# Summary:

The histogram displays the monthly passenger counts for the year 1960, which recorded the highest total number of air passengers. The chart also highlights clear seasonal trends, with the summer months of June, July, and August showing significantly higher passenger numbers compared to other months. This peak in the middle of the year likely corresponds to the holiday travel season or summer specifically. For instance, July had an approximate count of 600+ passengers, the highest for the year, reflecting the peak travel season.

Conversely, the months of January, February, and November show relatively lower passenger counts, with November showing the least with passengers less than 400, the lowest in the year. This pattern is typical as fewer people travel during the winter months, Overall, the histogram effectively illustrates the cyclical nature of air travel demand.
